### 2020-09-08 paper/main.tex
Starting to write

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage[margin=2cm]{geometry}
\usepackage{times}

\usepackage{lineno}

\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,      
    urlcolor=cyan,
}

\title{OvRF review}
\author{Laura Mu\~noz Baena and Art Poon}
\date{2020}

\begin{document}

\maketitle

\pagewiselinenumbers

\section {Introduction}
Overlapping genes are nucleotide sequences that encode more than one protein. 
There is a general consensus on the reasons to explain overlapping genes in viruses. 
The compression idea states that this type of gene arrangement allow viruses to maximize the amount of information contained in their genomes\cite{lamb1991diversity, pavesi1997informational}. 
This hypothesis is consistent with the idea that viral genomes tend to be small in order to prevent deleterious mutations and carry less replication times\cite{belshaw2008pacing}, improving their infectivity. However, the discovery of overlapping genes in longer genomes including both prokaryotic\cite{normark1983overlapping, johnson2004properties, palleja2008large} and eukaryotic organisms\cite{spencer1986overlapping, williams1986mouse, makalowska2005overlapping} challenges the compression theory.
A second theory suggests that overlapping genes could have potential regulatory roles\cite{normark1983overlapping, kapranov2007genome}.
This was first suggested by Normak and collaborators\cite{normark1983overlapping}, which examined overlapping genes in \textit{Escherichia coli}, and described the possible implications in regulation of gene expression. 
In viruses, the effects of overlapping genes in transcription and translation, have been studied for example in the Autographa californica nuclear polyhedrosis virus (AcNPV)\cite{friesen1985temporal}, the Bacteriohage MS2\cite{scherbakov2000overlapping} and Simian virus 40(SV40)\cite{buchman1984complex}.
The idea that overlapping regions could be used by viruses to create \textit{de novo} genes by a process known as "overprinting" was first suggested by Grasse\cite{grasse2013evolution} in 1973.
Even though several papers on this topic have been written since then, the core ideas regarding overprinting process remain the same: 
1. In eukaryotes, genes product of overprinting are duplicated and separated in different sections of the genome, but in viruses genes tend to remain overlapped due to size constrains. 
2. Genes that comes from overprinting have an unusual codon usage and encode proteins that have diverse functions specialized to the current life-style of the organism in which they are found. 
More recent studies have shown that overlapping regions tend to have higher levels of intrinsic structural disorder, likely because one of the members was recently born \textit{de novo}\cite{willis2018gene}.
The way overlapping genes arrange vary according to every organism. However, there are five frame shift possibilities for each overlap in reference with the parental gene (which we refer as +0 reading frame).
If the genes are in the positive strand, there can be genes with a frame shift of +1 or +2.
If the genes are in the negative strand, there can be genes in -0 (opposite strand without shifting), -1 or -2 reading frames.

\section{Materials and Methods}
\subsection{Overview of overlapping reading frames}
Initially, we downloaded an accession list of all viral genomes from NCBI \url{https://www.ncbi.nlm.nih.gov/genome/viruses/} that has information regarding: representative, neighbor, host, taxonomy and segment name for every virus entry on the database. 
% Re-making this part of the analysis. Will have to updated scripts and methods
We then used a Python script to complement information of the entries to include taxonomical information, genome lengt and total number of proteins. 
The same script was used to download all of the proteins associated with each accession number and their position on the genome. 
We used the protein file to find overlapping proteins and calculate the frameshift and number of nucleotides involved in each overlap/ 
We also created a script to classify each viral family according to the Baltimore classification displayed on ViralZone \url{https://viralzone.expasy.org/}.
The final result is a data base with a total of 11,891 viral entries, from which 9,928 are classified in any Baltimore groups. 
From remaining 1,963 entries, we can say, according to the information obtained from NCBI, that 400 are DNA viruses and 1,552 are RNA.

\subsection{Adjacency graphs}
To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family all the reference complete genomes from \url{https://www.ncbi.nlm.nih.gov/labs/virus/vssi/#/virus?SeqType_s=Nucleotide&SourceDB_s=RefSeq} as a fasta file. 
Over that file, we run a Python script to measure kmer-distance to create a distance matrix.
We then used that matrix as an input to form clusters representing similar proteins. 
Once we have each protein classified in a specific cluster, we generate a network plot, where each node represents a cluster, and the edges between clusters are formed when two proteins are adjacent to each other.

\newpage
\bibliography{main}
\bibliographystyle{vancouver}

\end{document}

### 2020-09-14 paper/main.tex
Working on methods section

@@
 \section{Materials and Methods}
-\subsection{Overview of overlapping reading frames}
-Initially, we downloaded an accession list of all viral genomes from NCBI \url{https://www.ncbi.nlm.nih.gov/genome/viruses/} that has information regarding: representative, neighbor, host, taxonomy and segment name for every virus entry on the database.
-% Re-making this part of the analysis. Will have to updated scripts and methods
-We then used a Python script to complement information of the entries to include taxonomical information, genome lengt and total number of proteins. 
-The same script was used to download all of the proteins associated with each accession number and their position on the genome. 
-We used the protein file to find overlapping proteins and calculate the frameshift and number of nucleotides involved in each overlap/ 
+\subsection{Data collection and processing}
+The initial part of the analysis consisted on the creation of the database.
+A list of all virus genomes was downloaded from NCBI \url{https://www.ncbi.nlm.nih.gov/genome/viruses/}.
+The table had information regarding: representative, neighbor, host, taxonomy and segment name for every reference genome on the database. 
+The neighbor information is a new approach from NCBI designed to include more diversity for every species, so every reference genome have some other genomes associated to it that are also complete or near complete.
+However, for the purpose of this study, we considered that the reference genome of each species is sufficient information. 
+Information of only the reference genomes was recorded and complemented by pulling data from the GeneBank database including: genome length, number of proteins, topology and molecule type using a Python script called scraper-2020.py.
+The same script was used to create a different table in which we stored all the CDSs of every genome with accession number (of the genome they belong to), product, strand, coordinates and start codon position.
+A second script (find$_$ovrfs.py) was used to group rows of the CDSs table by accession number and compare the coordinates of each protein.
+The output is a table with accession number of the reference genome, product, location and length of the overlapping proteins, number of nucleotides involved in the overlap and frame shift.
+%Update this information with new downloads
 We also created a script to classify each viral family according to the Baltimore classification displayed on ViralZone \url{https://viralzone.expasy.org/}.
@@
 
-\subsection{Adjacency graphs}
-To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family all the reference complete genomes from \url{https://www.ncbi.nlm.nih.gov/labs/virus/vssi/#/virus?SeqType_s=Nucleotide&SourceDB_s=RefSeq} as a fasta file.
-Over that file, we run a Python script to measure kmer-distance to create a distance matrix.
-We then used that matrix as an input to form clusters representing similar proteins. 
-Once we have each protein classified in a specific cluster, we generate a network plot, where each node represents a cluster, and the edges between clusters are formed when two proteins are adjacent to each other.
+\subsection{Overview of overlapping reading frames in virus genomes}
+
+\subsection{Clustering protein data}
+To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family all CDSs of the reference genomes from \url{https://www.ncbi.nlm.nih.gov/labs/virus/vssi/#/virus?SeqType_s=Nucleotide&SourceDB_s=RefSeq}.
+The information for all CDSs was stored in a multi-fasta file where each entry consisted in a header with the name of the virus, the accession number of the genome, the product name of the CDS, the strand in which it is encoded, the coordinates in the genome and the aminoacid sequence.
+Because aligned-based methods are computationally expensive, specially for highly divergent sequences, we used a k-mer frequency method to compare the aminoacid sequences of virus families from different Baltimore groups.
+The distance matrix generated from the k-mer distance measure, is used as an input for an R script that uses a clustering method to clasify proteins in clusters according to their similarities. 
+% Describe clustering method
+
+\subsection{Creation of adjacency graphs}
+A Python script was created to automatically generate adjacency plots to represent the organization of viral genomes across the family. 
+The script use as an input a csv file in which each line is a protein with information regarding the genome it belongs to, the coordinates where it is located and the cluster it belongs to. 
+We use the information to find the order of the proteins for each genome and labeled them depending on if they overlap or not. 
+Using Graphviz (Graph Visualization Software) \url{https://graphviz.org/theory/} to create a dot file of a network graph where each node represents a cluster, and each edge represent adjacent proteins between two clusters.
+Overlapping proteins are colored in blue, weather adjacent proteins are colored in gray. 
+As complementary information, the script uses Matplotlib\cite{Hunter:2007} to generate maps of each viral genome (where each protein is colored according to their cluster classification), and a wordcloud to visualize the names of the proteins in each cluster.
 

### 2020-09-22 scripts/new_viz_ovrf.py
Small fix to improve arrow look

@@ -304,6 +304,7 @@ def main():
     for cluster in cluster_list:
         cluster_size = len(cluster.proteins)
         node_size = math.sqrt(cluster_size)/3
+        edge_length = str(8)
 
         # Create a node
         dot.node(cluster.cluster, label=None, fixedsize="true", width=str(node_size), height=str(node_size),
@@ -314,7 +315,7 @@ def main():
         for adj_cluster, count in cluster.adjacent_clust.items():
             if count >= min_edge:
                 dot.edge(cluster.cluster, adj_cluster, label = None, penwidth = str(count),
-                color = "grey76", arrowsize = str(0.01), len = str(10))
+                color = "grey76", arrowsize = str(1/count), len = edge_length)
 
         # print("Overlapping", cluster.overlapping_clust)
         # overlap_count = 0
@@ -322,18 +323,18 @@ def main():
         for overlap_cluster, count in cluster.overlapping_clust.items():
             if count >= min_edge:
                 dot.edge(cluster.cluster, overlap_cluster, label = None, penwidth=str(count),
-                color="#143D59", arrowsize = str(0.01), len = str(10))
+                color="#143D59", arrowsize = str(1/count), len = edge_length)
                 if cluster.cluster == overlap_cluster:
                     print(f"Cluster: {cluster.cluster}, Numer of self edges: {count}")
 
     dot.render(filename=f"{args.outfile}.dot")
 
-    # Create wordcloud plot
-    wordcloud = wordcloud_plot(cluster_list)
-    wordcloud.savefig(f'wordcloud_{args.outfile}.pdf')
-    # Create genome plot
-    gen_plot = genome_plot(colors, genome_list)
-    gen_plot.savefig(f'genome_{args.outfile}.pdf')
+    # # Create wordcloud plot
+    # wordcloud = wordcloud_plot(cluster_list)
+    # wordcloud.savefig(f'wordcloud_{args.outfile}.pdf')
+    # # Create genome plot
+    # gen_plot = genome_plot(colors, genome_list)
+    # gen_plot.savefig(f'genome_{args.outfile}.pdf')
 
 
 if __name__ =='__main__':

### 2020-09-23 paper/main.tex
Working on paper, small fix to visualization

@@
 From remaining 1,963 entries, we can say, according to the information obtained from NCBI, that 400 are DNA viruses and 1,552 are RNA. 
-
-\subsection{Overview of overlapping reading frames in virus genomes}
+R programming language was used to generate the plots displaying the distribution of the database.
 
@@
 Because aligned-based methods are computationally expensive, specially for highly divergent sequences, we used a k-mer frequency method to compare the aminoacid sequences of virus families from different Baltimore groups.
-The distance matrix generated from the k-mer distance measure, is used as an input for an R script that uses a clustering method to clasify proteins in clusters according to their similarities. 
-% Describe clustering method
+The k-mer distance  matrix was used as input for a t-SNE (t-distributed stochastic neighbor embedding) algorithm using the Rtsne package on R. 
+With the t-SNE algorithm we grouped similar proteins that have a high likelihood of being together in a high dimensional space and then used a then used hclust to get the hierarchical clustering of the proteins under the method "ward.D2".
+We finally used cutree to divided our dendogram into the optimum number of clusters.
 
@@
 A Python script was created to automatically generate adjacency plots to represent the organization of viral genomes across the family. 
-The script use as an input a csv file in which each line is a protein with information regarding the genome it belongs to, the coordinates where it is located and the cluster it belongs to. 
-We use the information to find the order of the proteins for each genome and labeled them depending on if they overlap or not. 
+The script use as an input a csv file in which each line is a protein with information regarding the genome it belongs to, the coordinates where it is located and the cluster where it was assigned
+We use the information to find the order of the proteins for each genome and labeled them depending on whether they overlap or not. 
 Using Graphviz (Graph Visualization Software) \url{https://graphviz.org/theory/} to create a dot file of a network graph where each node represents a cluster, and each edge represent adjacent proteins between two clusters.
-Overlapping proteins are colored in blue, weather adjacent proteins are colored in gray. 
+Edges between Overlapping proteins are colored in blue and adjacent proteins are colored in gray. 
 As complementary information, the script uses Matplotlib\cite{Hunter:2007} to generate maps of each viral genome (where each protein is colored according to their cluster classification), and a wordcloud to visualize the names of the proteins in each cluster.
 
+\section{Results}
+\subsection{Overview of Overlapping Reading Frames in virus genomes}
+\begin{itemize}
+    \item Number of OvRFs vs Genome length
+    \item Overlap length vs Genome length
+    \item Frame shifts vs Baltimore classification
+\end{itemize}
+
+\subsection{Visualization of Overlapping Reading Frames across virus families}
 \newpage

### 2020-09-23 scripts/new_viz_ovrf.py

@@ -329,12 +329,12 @@ def main():
 
     dot.render(filename=f"{args.outfile}.dot")
 
-    # # Create wordcloud plot
-    # wordcloud = wordcloud_plot(cluster_list)
-    # wordcloud.savefig(f'wordcloud_{args.outfile}.pdf')
-    # # Create genome plot
-    # gen_plot = genome_plot(colors, genome_list)
-    # gen_plot.savefig(f'genome_{args.outfile}.pdf')
+    # Create wordcloud plot
+    wordcloud = wordcloud_plot(cluster_list)
+    wordcloud.savefig(f'wordcloud_{args.outfile}.pdf')
+    # Create genome plot
+    gen_plot = genome_plot(colors, genome_list)
+    gen_plot.savefig(f'genome_{args.outfile}.pdf')
 
 
 if __name__ =='__main__':

### 2020-09-28 scripts/find_ovrfs.py
Scripts on Nagisa, preparing for merge

@@ -59,7 +59,7 @@ def find_ovrfs(accn, records, outfile):
                 for l2, r2 in c2:
                     left = max(l1, l2)
                     right = min(r1, r2)
-                    overlap = right - left
+                    overlap = (right - left) +1
 
                     shift = abs(xl1-xl2) % 3
                     if dir1 != dir2:
@@ -74,11 +74,10 @@ def find_ovrfs(accn, records, outfile):
                         ))
 
 
-outfile = open('../data/find_ovrfs.csv', 'w')
+outfile = open('../data/find_ovrfs2.csv', 'w')
 outfile.write('accn,prod1,loc1,dir1,prod2,loc2,dir2,seqlen1,seqlen2,overlap,shift\n')
 
 for accn, records in get_records('../data/orfs-fixed.csv'):
     #print(accn)
     find_ovrfs(accn, records, outfile)
-
 

### 2020-09-28 scripts/reduce-overlaps.py

@@ -4,14 +4,14 @@ occur in multiple segments because of "introns".
 """
 
 from csv import DictReader, DictWriter
-handle = open('../data/find_ovrfs.csv')
+handle = open('../data/find_ovrfs2.csv')
 reader = DictReader(handle)
 
 last_state = None
 last_overlap = 0
 last_shift = None
 
-outfile = open('../data/ovrfs-reduced.csv', 'w')
+outfile = open('../data/ovrfs-reduced2.csv', 'w')
 writer = DictWriter(outfile, fieldnames=[
     'accn', 'prod1', 'loc1', 'dir1', 'prod2', 'loc2', 'dir2', 'seqlen1', 'seqlen2', 'overlap', 'shift'
 ])

### 2020-09-28 scripts/scraper-edited.py

@@ -2,7 +2,7 @@ from Bio import Entrez, SeqIO, SeqFeature
 from csv import DictWriter
 from time import sleep
 
-Entrez.email = 'bioinfo@uwo.ca'
+Entrez.email = 'lmuoz@uwo.ca'
 #'bioinfo@uwo.ca'
 #accn = 'NC_015932'
 tbl = '../data/taxid10239.tbl'
@@ -103,9 +103,6 @@ def retrieve_annotations(record):
 
 def main():
 
-    #orffile = open('orfs.csv', 'w')
-    #orffile.write('accno,product,strand,coords,aaseq\n')
-
     handle2 = open('../data/virus_with_info.csv', 'w')
     virus_info_file = DictWriter(handle2,
                                      delimiter = ',',
@@ -125,15 +122,21 @@ def main():
         else:
             accnos = row['Accession']
 
-        gid = retrieve_gid(accnos)
-        record = retrieve_record(gid)
-        annotation = retrieve_annotations(record)
-        final_row = {**row, **annotation}
-        virus_info_file.writerow(final_row)
         handle2.flush()
 
-        print(row['Genome'], gid)  # track progress
-        sleep(1)
+        for accn in accnos:
+            gid = retrieve_gid(accn)
+            if gid is None:
+                print('Warning, failed to retrieve gid for {}'.format(accn))
+                continue
+
+            print(row['Genome'], gid)  # track progress
+            record = retrieve_record(gid)
+            annotation = retrieve_annotations(record)
+            final_row = {**row, **annotation}
+            virus_info_file.writerow(final_row)
+
+            sleep(1)
 
 if __name__ == "__main__":
     main()

### 2020-09-28 scripts/scraper.py

@@ -97,18 +97,18 @@ def retrieve_CDS(record):
 
 
 def main():
-    handle = open('../data/viruses.csv', 'w')
-    outfile = DictWriter(handle,
-                        delimiter=',',
-                        extrasaction='ignore',
-                        fieldnames=[
-                            'Family', 'Genome', 'Source information',
-                            'Accession', 'Date completed', 'Date updated',
-                            'Genome length', 'Number of proteins', 'Host'
-                        ])
-    outfile.writeheader()
+    # handle = open('../data/viruses.csv', 'w')
+    # outfile = DictWriter(handle,
+    #                     delimiter=',',
+    #                     extrasaction='ignore',
+    #                     fieldnames=[
+    #                         'Family', 'Genome', 'Source information',
+    #                         'Accession', 'Date completed', 'Date updated',
+    #                         'Genome length', 'Number of proteins', 'Host'
+    #                     ])
+    # outfile.writeheader()
 
-    orffile = open('orfs.csv', 'w')
+    orffile = open('orfs_with_nt.csv', 'w')
     orffile.write('accno,product,strand,coords,aaseq\n')
 
     for row in parse_table(tbl):
@@ -118,8 +118,8 @@ def main():
         else:
             accnos = row['Accession']
 
-        outfile.writerow(row)
-        handle.flush()
+        # outfile.writerow(row)
+        # handle.flush()
 
         for accn in accnos:
             gid = retrieve_gid(accn)

### 2020-09-28 scripts/old_scripts/find_ovrfs.py
Re calculating ovrfs for new database

@@ -30,6 +30,7 @@ def get_records(file):
 def parse_coords(coords):
     res = []
     for interval in coords.split(';'):
+        print(interval)
         left, right = interval.split(':')
         res.append((int(left), int(right)))
     return(res)
@@ -40,9 +41,11 @@ def find_ovrfs(accn, records, outfile):
     :param records:
     :return: list of overlapping regions, if any
     """
+    print(accn)
     n = len(records)
     for i in range(n):
         rec1 = records[i]
+        print(rec1['coords'], type(rec1['coords']), len(rec1))         
         c1 = parse_coords(rec1['coords'])
         xl1 = min([l for l, r in c1])  # extreme left
         len1 = sum([r-l for l, r in c1])  # total CDS length
@@ -66,7 +69,7 @@ def find_ovrfs(accn, records, outfile):
                         shift = '-'+str(shift)
                     else:
                         shift = '+'+str(shift)
- 
+
                     if overlap > 0:
                         outfile.write('{},"{}",{},{},"{}",{},{},{},{},{},{}\n'.format(
                             accn, rec1['product'], xl1, dir1, rec2['product'], xl2, dir2,
@@ -74,10 +77,9 @@ def find_ovrfs(accn, records, outfile):
                         ))
 
 
-outfile = open('../data/find_ovrfs2.csv', 'w')
+outfile = open('find_ovrfs_out.csv', 'w')
 outfile.write('accn,prod1,loc1,dir1,prod2,loc2,dir2,seqlen1,seqlen2,overlap,shift\n')
 
-for accn, records in get_records('../data/orfs-fixed.csv'):
+for accn, records in get_records('/home/lmunoz/Projects/ovrf-review/dataset_2020/total_orfs.csv'):
     #print(accn)
     find_ovrfs(accn, records, outfile)
-

### 2020-09-30 paper/main.tex
Updating database info, cluster formation

@@
 A list of all virus genomes was downloaded from NCBI \url{https://www.ncbi.nlm.nih.gov/genome/viruses/}.
-The table had information regarding: representative, neighbor, host, taxonomy and segment name for every reference genome on the database. 
+The table of 247,941 lines, had information regarding: representative, neighbor, host, taxonomy and segment name for every reference genome on the database. 
 The neighbor information is a new approach from NCBI designed to include more diversity for every species, so every reference genome have some other genomes associated to it that are also complete or near complete.
-However, for the purpose of this study, we considered that the reference genome of each species is sufficient information. 
-Information of only the reference genomes was recorded and complemented by pulling data from the GeneBank database including: genome length, number of proteins, topology and molecule type using a Python script called scraper-2020.py.
-The same script was used to create a different table in which we stored all the CDSs of every genome with accession number (of the genome they belong to), product, strand, coordinates and start codon position.
-A second script (find$_$ovrfs.py) was used to group rows of the CDSs table by accession number and compare the coordinates of each protein.
-The output is a table with accession number of the reference genome, product, location and length of the overlapping proteins, number of nucleotides involved in the overlap and frame shift.
+However, for the purpose of this study, we considered the reference genome of each species as sufficient information. 
+We used a Python script to retrieve relevant information about each reference genome by pulling data from the GeneBank database. 
+The new table included the following complementary information: genome length, number of proteins, topology and molecule type.
+The same script was used to create a different table in which we stored information for all the CDSs of every genome with accession number (of the genome they belong to), product, strand, coordinates and start codon position.
+A second script (find$_$ovrfs.py) was used to group rows of the CDSs information table by accession number and compare the coordinates of each protein in order to obtain the possible overlaps. 
+The output is a table with accession number of the reference genome, product, location and direction of each protein, number of nucleotides involved in the overlap and frame shift.
 %Update this information with new downloads
 We also created a script to classify each viral family according to the Baltimore classification displayed on ViralZone \url{https://viralzone.expasy.org/}.
-The final result is a data base with a total of 11,891 viral entries, from which 9,928 are classified in any Baltimore groups. 
-From remaining 1,963 entries, we can say, according to the information obtained from NCBI, that 400 are DNA viruses and 1,552 are RNA. 
+The result of the processing steps, is a table of 12,609 reference genomes from which 9,982 are classified in any Baltimore groups. 
+From the remaining 2,627, 1,001 are DNA viruses and 1,626 are RNA. according to the information obtained from NCBI.
+The second table contains the information of 160,222 overlaps from all the reference genomes.
 R programming language was used to generate the plots displaying the distribution of the database.
@@
 \subsection{Clustering protein data}
-To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family all CDSs of the reference genomes from \url{https://www.ncbi.nlm.nih.gov/labs/virus/vssi/#/virus?SeqType_s=Nucleotide&SourceDB_s=RefSeq}.
-The information for all CDSs was stored in a multi-fasta file where each entry consisted in a header with the name of the virus, the accession number of the genome, the product name of the CDS, the strand in which it is encoded, the coordinates in the genome and the aminoacid sequence.
-Because aligned-based methods are computationally expensive, specially for highly divergent sequences, we used a k-mer frequency method to compare the aminoacid sequences of virus families from different Baltimore groups.
-The k-mer distance  matrix was used as input for a t-SNE (t-distributed stochastic neighbor embedding) algorithm using the Rtsne package on R. 
-With the t-SNE algorithm we grouped similar proteins that have a high likelihood of being together in a high dimensional space and then used a then used hclust to get the hierarchical clustering of the proteins under the method "ward.D2".
+To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family the amino acid sequences for all CDSs of the reference genomes from the NCBI virus database.
+The header of each entry in the multi-fasta file has the the accession number of the genome, the product name of the CDS, the strand in which it is encoded and the coordinates in the genome.
+Since our analysis is between highly divergent sequences (entire virus families), we use \textit{k-}mer counts to compare the sequences. 
+The method projects each of the inpit seqences into a feature space of \textit{k-}mer counts where sequence information is transformed into numerical values that ae latter used to calculate distances between all possible sequence pairs in the dataset.
+The k-mer distance  matrix was used as input for a t-distributed Stochastic Neighbor Embedding algorithm (t-SNE)using the Rtsne package on R. 
+The purpose of t-SNE was to exaggerate the distance between proteins to  extract functioning clusters to work with by grouping similar proteins that have a high likelihood of being together in a high dimensional space.
+We then used hclust to get the hierarchical clustering of the proteins under the method "ward.D2".
 We finally used cutree to divided our dendogram into the optimum number of clusters.

### 2020-10-09 paper/main.tex
Paper: Progress on results section

@@
 The same script was used to create a different table in which we stored information for all the CDSs of every genome with accession number (of the genome they belong to), product, strand, coordinates and start codon position.
-A second script (find$_$ovrfs.py) was used to group rows of the CDSs information table by accession number and compare the coordinates of each protein in order to obtain the possible overlaps. 
+A second script (find$\_$ovrfs.py) was used to group rows of the CDSs information table by accession number and compare the coordinates of each protein in order to obtain the possible overlaps. 
 The output is a table with accession number of the reference genome, product, location and direction of each protein, number of nucleotides involved in the overlap and frame shift.
-%Update this information with new downloads
-We also created a script to classify each viral family according to the Baltimore classification displayed on ViralZone \url{https://viralzone.expasy.org/}.
-The result of the processing steps, is a table of 12,609 reference genomes from which 9,982 are classified in any Baltimore groups. 
-From the remaining 2,627, 1,001 are DNA viruses and 1,626 are RNA. according to the information obtained from NCBI.
-The second table contains the information of 160,222 overlaps from all the reference genomes.
+We also created a script to classify each viral family according to the Baltimore classes displayed on ViralZone \url{https://viralzone.expasy.org/}.
+The result of the processing steps, is a table of 12,609 reference genomes from which 9,982 are classified in any Baltimore groups and 6,614 have overlapping proteins. 
+From the remaining 2,627, 1,001 are DNA viruses and 1,626 are RNA according to the information obtained from NCBI.
+The second table contains information of 160,222 overlapping proteins from all the reference genomes.
 R programming language was used to generate the plots displaying the distribution of the database.
 
-\subsection{Clustering protein data}
-To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family the amino acid sequences for all CDSs of the reference genomes from the NCBI virus database.
+\subsection{Clustering protein data by family}
+To analyze the organization and distribution of overlapping reading frames in different virus families, we downloaded for each family protein sequences for all CDSs of the reference genomes from the NCBI virus database.
 The header of each entry in the multi-fasta file has the the accession number of the genome, the product name of the CDS, the strand in which it is encoded and the coordinates in the genome.
-Since our analysis is between highly divergent sequences (entire virus families), we use \textit{k-}mer counts to compare the sequences. 
-The method projects each of the inpit seqences into a feature space of \textit{k-}mer counts where sequence information is transformed into numerical values that ae latter used to calculate distances between all possible sequence pairs in the dataset.
-The k-mer distance  matrix was used as input for a t-distributed Stochastic Neighbor Embedding algorithm (t-SNE)using the Rtsne package on R. 
+Since our analysis is between highly divergent sequences (entire virus families), we use \textit{k-}mer counts to compare the amino acid sequences. 
+The method projects each of the input sequences into a feature space of \textit{k-}mer counts where sequence information is transformed into numerical values that are latter used to calculate distances between all possible sequence pairs in the dataset\cite{zielezinski2019benchmarking}.
+The k-mer distance  matrix was used as input for a t-distributed Stochastic Neighbor Embedding algorithm (t-SNE) using the Rtsne package on R. 
 The purpose of t-SNE was to exaggerate the distance between proteins to  extract functioning clusters to work with by grouping similar proteins that have a high likelihood of being together in a high dimensional space.
@@
 We use the information to find the order of the proteins for each genome and labeled them depending on whether they overlap or not. 
-Using Graphviz (Graph Visualization Software) \url{https://graphviz.org/theory/} to create a dot file of a network graph where each node represents a cluster, and each edge represent adjacent proteins between two clusters.
+With Graphviz (Graph Visualization Software) \url{https://graphviz.org/theory/}, we produced a dot file of a network graph where each node represents a cluster, and each edge represent adjacent proteins between two clusters.
 Edges between Overlapping proteins are colored in blue and adjacent proteins are colored in gray. 
-As complementary information, the script uses Matplotlib\cite{Hunter:2007} to generate maps of each viral genome (where each protein is colored according to their cluster classification), and a wordcloud to visualize the names of the proteins in each cluster.
+As complementary information, the script uses Matplotlib\cite{Hunter:2007} to generate maps of each viral genome (where each protein is colored according to their cluster classification), and WordClouds to visualize the names of the proteins in each cluster.
 
 \section{Results}
-\subsection{Overview of Overlapping Reading Frames in virus genomes}
-\begin{itemize}
-    \item Number of OvRFs vs Genome length
-    \item Overlap length vs Genome length
-    \item Frame shifts vs Baltimore classification
-\end{itemize}
+\subsection{Number of ORFs is correlated with number of OvRFs}
+As expected, there is a positive correlation between the number of ORFs and the number of proteins overlapping (Person's product-moment correlation, $P<10^{-12}$).
+However, there appear to be different tendencies for different Baltimore groups. 
+The biggest group of viruses, double stranded DNA viruses, encode on average 202 proteins and 37 overlaps. 
+Some extreme cases of this group are  Paramecium bursaria Chlorella viruses (which host is eykaryotic algae)from the Phycodnaviridae family, have around 1650 ORFS and  more than 600 overlaps.
+37\% of its proteins have an overlap. 
+%The \textit{Siphoviridae} family, that represents 42\% of that baltimore group. 
+In contrast, positive single stranded RNA, encode on average 10.5 from which around 3 are involved in an overlap. 
+Negative single stranded RNA encode around 7 proteins and have 2 overlaps, that has some extreme examples like Simian hemorrhagic fever virus (NC$\_003092$), that encodes 33 proteins from which 51\% are involved in an overlap.
+
+\subsection{Genome length is not correlated with overlap length}
+We face a completely different scenario when we evaluate the correlation between genome length and mean overlap length.
+There are no clear trends in this case, and viruses are divided across the entire spectrum between 1 and 10,000 nucleotides involved in overlaps for viruses lengths varying from 1 thousand to 1 million nucleotides long.
+As we pointed out before, the longest genomes are those of DNA viruses, but the longer overlaps, are from RNA viruses like
+
+\subsection{Frame Shifts distribution}
+Different Baltimore groups also tend to have different kinds of overlaps. 
+The +2 and +1 frame shifts are common for all Baltimore groups, being more common in double stranded and single stranded viruses respectively. 
+The only gruops with negative sense frame shifts are negative ssRNA and DNA viruses (both single and double stranded).
 

### 2021-12-08 dataset_2020/df_inspection.R
adding random noise, cheking bad entries

@@ -1,9 +1,15 @@
 library(dplyr)
+library(treemap)
+library(plyr)
+library(ggplot2)
+library(ggridges)
+require(grid)
 ########################################################
 # Plot frame shift distribution
 ########################################################
 setwd('/home/lmunoz/Projects/ovrf-review/dataset_2020/')
-overlaps <- read.csv('2_overlapping_march2020.csv', colClasses='character')
+#overlaps <- read.csv('2_overlapping_march2020.csv', colClasses='character')
+overlaps <- read.csv('overlaps_with_noise.csv', colClasses='character')
 
 overlaps$extreme_left1 <- as.integer(overlaps$extreme_left1)
 overlaps$extreme_left2 <- as.integer(overlaps$extreme_left2)
@@ -15,9 +21,12 @@ overlaps$seqlen2 <- as.integer(overlaps$seqlen2)
 overlaps$overlap <- as.integer(overlaps$overlap)
 overlaps$shift <- factor(overlaps$shift, 
                          levels=c('-2', '-1', '-0', '+0', '+1', '+2'))
+non_filtered_ovps<- overlaps
 
+### Find duplicates with different overlap lengths and merge them (remove 1426 entries)
+overlaps2 <- overlaps %>% group_by(across(c(-overlap))) %>% summarise(overlap = sum(overlap))
 
-# Remove entries where number of nucleotides is not consequent with frame shift
+# Remove 5241 entries where number of nucleotides is not consequent with frame shift
 non_splicing_overlaps <- overlaps %>%
   filter(
     ((shift == "+0" | shift == "-0") & overlap%%3 == 0) |
@@ -31,6 +40,7 @@ circular <- which(non_splicing_overlaps$extreme_left1==non_splicing_overlaps$ext
 
 non_splicing_overlaps <- non_splicing_overlaps[-circular,]
 
+overlaps <- non_splicing_overlaps
 
 # Ridgeplot
 plot <- ggplot(non_splicing_overlaps, aes(x = log10(overlap), y = shift, group = shift, fill = shift)) +
@@ -43,9 +53,11 @@ plot
 #palete<-c("#32621f", "#769845", "#cacfa6", "#f1dba2", "#de8f42", "#c1462b")
 # +2, +1, +0, +1, -1, -2
 
-library(treemap)
-library(plyr)
-require(grid)
+# Introduce Noise: Randomly select rows and change coordinates 
+n<-3
[…]

### 2021-12-08 paper/response.md
added response Markdown document

# Major Issues

## Reviewer 1

1. Annotation reliability

>The authors rely on NCBI annotations of CDSs. However, these tend to be quite inaccurate (as the authors discovered for bovine adenovirus NC 002513, p20 lines 5-8). Overlapping and other non-standard genes (ribosomal frameshifting, non-AUG initiation, transcriptional slippage, alternative splicing, misannotation of an upstream in-frame AUG when in fact it is not actually used due to being upstream of a transcript start site, etc) are common in viruses but often dealt with poorly by automatic gene prediction software or indeed by manual annotators. Thus, many virus sequences have missing ORFs, while some sequences may have spurious ORFs annotated as if they were real CDSs.
>A few misannotations could seriously bias results - particularly for reverse-frame overlapping genes - for example, the authors 4 reverse-frame overlaps in dsRNA viruses arise from "NC\_003729" Reoviridae, "NC\_042073" Cystoviridae, "NC\_043677" Chrysoviridae, "NC\_043678" Chrysoviridae and are likely all misannotations.
>The authors should carefully curate their input dataset and/or show that their methods and conclusions are robust with respect to misannotations.

Thank you for raising this issue.  It is a true and unfortunate fact that many records in Genbank are misannotated, even for entries in the RefSeq database that have presumably been manually curated.  Unfortunately, due to the scale of our analysis (all available virus genomes), it was not feasible to curate the entire data set of 12,609 genomes (451,228 annotated reading frames) in the time allotted to revise our manuscript.  Therefore, we followed the reviewer's suggestion by focusing our manual curation on specific cases, and by  evaluating how sensitive our methods and conclusions are to misannotation in the database.

First, we had screened for misannotations by testing whether the length of overlapping regions was consistent with apparent frameshift of the respective reading frames.  For example, an overlap between reading frames that are shifted by `+2` should have a length that is not divisible by 3, with remainder 1, *i.e.*, length mod 3 = 1.  Applying this filter to our data identified 5,241 entries with misannotations, which we discarded from further analysis.  Due to space constraints, this quality control step was not described in our original manuscript.  We apologize for this omission and have added this explanation to the revision.

reading frames associated with each overlap comprised the expected number of nucleotides

We are aware of the inconvenient that would rise from missannotation problems.  Performing analysis on a broad scale as intended on this research is possible by using annotated genomes from a curated dataset, as we would expect the reference genomes from NCBI to be.  However, in order to consolidate our results as robust despite misannotations, we have performed the following filtration steps:

* In order to ensure reliability on the calculation of the frame shifts, we tested whether the frame shifts assigned to each overlap corresponded with the number of nucleotides expected in each case.
For example, you would expect the mode 3 of an overlap located in a +2 frame shift to be equal to one (`overlap%%3 == 1`).
Calculating the mode for all overlaps in our database lead us to remove 5,241 entries that did not follow such criteria.
* Alternatively, we created a python script to introduce noise in 10\% of our database.
The scripts selects random lines in total\_orfs.csv, the csv file containing all CDS from all viruses in our database, and samples a random number between 0 and the coordinates for both, the start and end locations of each CDS.
From the modified database, we re-calculated the overlapping regions and performed the same procedure using the scripts `df_inspection.R` and `plots_for_paper_2021.R`.

![](overlap_lengtht_noise.pdf)
Associations between overlap lengths and frame shifts after introduction of random noise to 10\% of the entries of the database

![](treemap_noise.pdf)
Distribution of frameshits across Baltimore classes with the modified database

2. Overlaps in the $+0$ frame shift

> The authors include +0 frame overlaps. By this, they mean where the same coding sequence forms part of two different CDS annotations, e.g. in Venezuelan equine encephalitis alphavirus NC\_001449, polyprotein P123 is encoded in the CDS 44-5684 and polyprotein P1234 is encoded in the CDS 44-7526 with readthrough of the stop codon at position 5682-5684. The authors count the region 44-5684 as a +0 frame overlap with the region 44-7526. The authors justification for counting this as an overlap is "Even though the reading frames share codons, they yield different gene products such that the codons are exposed to different selective environments." and "+0 overlaps increase the selective burden of the same nucleotides" (p19 lines 6-10). I find this difficult to swallow. First, the shared domains often have largely similar functions in the two products (e.g. alphavirus P123 versus the P123 domains of P1234) - cf. different-frame overlaps such as orthoreovirus sigma-1 and sigma-1s - where the two proteins have no sequence in common and hence no shared function at all. Second, all codons are subject to multiple selective pressures anyway (translation, mutational, protein function - many proteins are multifunctional); it doesn't make sense to me to single out +0 overlaps as being subject to additional selection pressures.

* Remove this type of overlaps from main plots, put the rest as supplementary material
* Remove lines from p19 lines 6-10

3. Biases on $+0$ frame shift

> A second problem with including +0 frame overlaps is that their presense in a genome is very much subject to the whim of the annotator. As an example, the alphavirus NC\_001449 has both readthrough and frameshift ORFs annotated:
> 45-5684 and 45-7526 (stop codon readthrough at 5682-5684)
7562-11329 and join(7562-9970,9970-10047) (ribosomal frameshift at 9970)
> leading to $+0$ frame overlaps 45-5684 and 7562-9970. On the other hand, the alphavirus NC\_023812 only has two CDSs annotated 43-7467 and 7541..11269. The stop codon read through and frameshift are there, but extra CDSs are not annotated so the authors find zero +0 frame overlaps.
> Such annotation issues will bias the results as these mainly affect RNA viruses.

We agree that this is a significant limitation in the analysis.
We expect that the results obtained by the noise introduction would demonstrate that even though such cases are inevitable, the overall trend of our results would be consistent given the large amount of analysed data.

4. Annotation bug causing wrong overlap measurement

> There seem to be issues with the annotation scripts. For example in the authors' supplementary overlaps.csv file, we have the following entries for NC\_001449, listing the +0 frame P123/P1234 overlap in the first line, but then the frameshift ORF join(7562-9970,9970-10047), comprising a +0 frame overlap 7562-9970 and a +2 frame overlap 9970-10047, is listed as two +0 frame overlaps:
> "NC\_001449","non-structural polyprotein precursor P123",44,5684,1,"non-structural polyprotein precursor P1234",44,7526,1,5640,7482,5640,+0 \\
"NC\_001449","truncated polyprotein",7561,10047,1,"structural polyprotein precursor",7561,11329,1,2487,3768,2409,+0 \\
"NC\_001449","truncated polyprotein",7561,10047,1,"structural polyprotein precursor",7561,11329,1,2487,3768,78,+0 \\
> Another example is in simian hemorrhagic fever virus NC\_003092, where the 678-nt +1 frame fragment (i.e. 2875-3552) of the join(210-2876,2875-3552) frameshift CDS is labelled as a +0 frame overlap.
"NC\_003092","ORF1aTF polyprotein",209,3552,1,"ORF1a polyprotein"MaSc,209,6527,1,3345,6318,2667,+0 \\
"NC\_003092","ORF1aTF polyprotein",209,3552,1,"ORF1a polyprotein",209,6527,1,3345,6318,678,+0 \\

This is an excellent observation.
This was the correct file. How did we deal with those cases? Removing overlaps product from splicing events.
The input for our calculation of overlaps is based on the file total\_orfs.csv that contains for each protein from each genome the accession number (accno), Product, Strand, Coordinates and Start Codon.
The misscalculation problem seems to be associated with entries that have internal frame shifts.
For example, for the record NC\_003092, the annotations for ORFS1ab protein and ORF1aTF polyprotein have some internal frame shifts that are not even spliced events: 209:6521;6520:10996 and 209:2876;2874:3552 respectively.
Similarly, for NC\_001449 the truncated polyprotein has coordinates: 7561:9970;9969:10047 (again, a frame shift of one nucleotide) where ribosome probably jumps back one nucleotide.
Calculating frame shift on such entries is challenging since there would be two different frame shifts for the same protein.

Following step: Detect how many entries we have with similar characteristics.

## Reviewer \#2

1. Consistency of alignment-free method
> In the graph analysis, homology is based on an alignment free method. how robust are the results of the graph analyses? Will similar conclusions be made with an alignment based homology method? How sensitive are the results and conclusions to the chosen cutoffs? To answer these questions it would be of value to repeat the graph analysis (at least on a small subset) with an alignment based method for homology clustering and possibly also modifying clustering cutoffs on the existing clustering and briefly discuss the impact of these changes on the results and conclusions of the analyses.

The creation of the adjacency graphs using an alignment-free method was motivated for previous difficulties that we encountered when aligning entire virus families. Such alignments tend to be affected for large periods of evolution that in viruses usually involve events such as recombination, insertion and deletions, and the lack of entire genes in different species.
We tried to reproduce the analysis in a small family such as Papillomaviridae but were not able to create a proper alignment from which to estimate aligment scores that would resemble the kmer-counts that we obtained with the clustering method.

2. Overlap length information on graphs

> The graph analysis is lacking information of the overlap frame and overlap length, which are of great interest. Adding edges separate for frame and for length in the supplementary information would be of value (or at least adjusting edge width by a minimum of overlap length and discussing the impact on the resulting graph).

Color edges in a gradient according with length of overlap

## Minor Issues}

### Reviewer 1

1. Suggested references
> While the authors have referenced previous studies such as Brandes & Linial, Chirico et al, and Schlub & Holmes, I feel that they should also reference Rancurel et al PMID 19640978 and Pavesi et al PMID 30339683, who used carefully curated datasets of overlaps rather than relying on NCBI annotation.

Include them

2. Specify terms
> The authors frequently (e.g. p5 line 24, p5 line 25, etc) use the term "reading frame" when the term "open reading frame" (i.e. "ORF") would, I feel, be clearer. A genome has 6 (potential) reading frames (viz. +0,+1,+2,-0,-1,-2) but (for large genomes) can have 100s of ORFs.

change them

3. Negative strand

> On p6 lines 10-11, I feel it would be useful to state explicitly how (-)strand ORFs (where the biological 5' and 3' ends are in the opposite orientation with respect to the genome coordinates) are dealt with.

4. Coronaviridae genome

> In Fig 4, why are there only 11 clusters in the coronaviridae graph, given that SARS alone has 13 ORFs and there are additional coronavirus accessory genes present in other coronaviruses?

Proteins with high homology to certain clussters can be grouped in that specific cluster and won't form their individual cluster.

5. Simian hemorrhagic fever genome

> The authors say that simian hemorrhagic fever virus genome (NC 003092) "encodes 33 reading frames" (p9 line 12). I don't know where this number came from because NC\_003092 only has ~15 CDSs.

Double check it

### Reviewer 2

1. Figures discrepancy

> There seems to be a discrepancy between figure 1c and figure 2a - in overlap of frame +2 the majority of overlap lengths is 1 and 4, and overlap of +2 is the dominant in both dsDNA and dsRNA, however in Figure 1c there are no peaks for dsDNA with length 1 and 4, this does not make sense because there are more dsDNA than dsRNA in the analysis. This apparent discrepancy needs to be fixed or well explained in the text.

2. Overlaps on -0 frame shift

> Long overlaps with -0 frame are not explained - is this unique to a single edge between homology clusters or does this occur in several edges, if the latter is true this might have implications for the evolution of the genetic code.

3. Conserved overlaps

> Although most adjacent nodes don't show overlap, some large nodes have thick edges (e.g., nodes 6 and 7 in the Adenoviridae family) suggesting that in some cases adjacency of specific genes is accompanied by conserved overlap between the two genes - this should be discussed.

### 2021-12-08 paper/response.md
updated response md

@@
 Unfortunately, due to the scale of our analysis (all available virus genomes), it was not feasible to curate the entire data set of 12,609 genomes (451,228 annotated reading frames) in the time allotted to revise our manuscript.
-Therefore, we followed the reviewer's suggestion by focusing our manual curation on specific cases, and by  evaluating how sensitive our methods and conclusions are to misannotation in the database.
+However, there are some quality control measures that can be automated.
 
-First, we had screened for misannotations by testing whether the length of overlapping regions was consistent with apparent frameshift of the respective reading frames.
+For instance, we have been screening for misannotations by testing whether the length of overlapping regions was consistent with apparent frameshift of the respective reading frames.
 For example, an overlap between reading frames that are shifted by `+2` should have a length that is not divisible by 3, with remainder 1, *i.e.*, length mod 3 = 1.
-Applying this filter to our data identified 5,241 entries with misannotations, which we discarded from further analysis.
+Applying this filter to our data identified 5,241 (~1%) entries with misannotations - these entries were discarded from further analysis.
+We manually determined that these cases were frequently caused by undocumented splicing or slippage.
+
+Thus, the cases of misannotation that the reviewer astutely identified have already been filtered out of our analysis, though those entries remained in the source data.
 Due to space constraints, this quality control step was not described in our original manuscript.
-We apologize for this omission and have added this explanation to the revision.
+We sincerely apologize for the confusion, and we have added this explanation to the revision.
 
-reading frames associated with each overlap comprised the expected number of nucleotides
-
-We are aware of the inconvenient that would rise from missannotation problems.
-Performing analysis on a broad scale as intended on this research is possible by using annotated genomes from a curated dataset, as we would expect the reference genomes from NCBI to be.
-However, in order to consolidate our results as robust despite misannotations, we have performed the following filtration steps:
-
-* In order to ensure reliability on the calculation of the frame shifts, we tested whether the frame shifts assigned to each overlap corresponded with the number of nucleotides expected in each case.
-For example, you would expect the mode 3 of an overlap located in a +2 frame shift to be equal to one (`overlap%%3 == 1`).
-Calculating the mode for all overlaps in our database lead us to remove 5,241 entries that did not follow such criteria.
-* Alternatively, we created a python script to introduce noise in 10\% of our database.
-The scripts selects random lines in total\_orfs.csv, the csv file containing all CDS from all viruses in our database, and samples a random number between 0 and the coordinates for both, the start and end locations of each CDS.
-From the modified database, we re-calculated the overlapping regions and performed the same procedure using the scripts `df_inspection.R` and `plots_for_paper_2021.R`.
+To further evaluate the sensitivity of our results to misannotation, we have performed a simulation experiment in which a random sample of reading frame coordinates were modified.
+Specifically, we selected a random sample of 5% of entries in the CSV file containing all coding sequences for viruses in the database, and then shifted the start and end coordinates by an amount drawn from a uniform distribution from -10 to 10.
+Next, we re-ran the analysis pipeline on the modified data to assess the impact of this error variation on our results.
 
@@
 
-* Remove this type of overlaps from main plots, put the rest as supplementary material
-* Remove lines from p19 lines 6-10
+We concede that sequences belonging to this +0 category of overlaps do not necessarily result in gene products that experience different selective environments.
+Hence, we have removed this justification from the manuscript, as well as the corresponding results from figures and text.
+We submit, however, that these cases should not be entirely disregarded.
+For instance, they represent a convenient control group --- the same sequence being used to produce different gene products, but in the same reading frame --- that provides context for interpreting results for standard overlapping reading frames.
+Therefore, we are providing supplementary material that includes the +0 cases so that this information is available, with the caveats raised by the reviewer.
 
@@
 
-> A second problem with including +0 frame overlaps is that their presense in a genome is very much subject to the whim of the annotator.
+> A second problem with including +0 frame overlaps is that their presence in a genome is very much subject to the whim of the annotator.
 As an example, the alphavirus NC\_001449 has both readthrough and frameshift ORFs annotated:
@@
 
-We agree that this is a significant limitation in the analysis.
-We expect that the results obtained by the noise introduction would demonstrate that even though such cases are inevitable, the overall trend of our results would be consistent given the large amount of analysed data.
+We agree that this is a significant limitation in the analysis (see above, removed +0 from results)
 
@@
 
-This is an excellent observation.
+Thank you for raising this issue.
+As noted in our response to an earlier point, cases that involved splices such as the example provided by the reviewer were excluded from our analysis.
+
 This was the correct file. How did we deal with those cases? Removing overlaps product from splicing events.
@@
 
-## Reviewer \#2
+## Reviewer #2
 
@@
 Such alignments tend to be affected for large periods of evolution that in viruses usually involve events such as recombination, insertion and deletions, and the lack of entire genes in different species.
-We tried to reproduce the analysis in a small family such as Papillomaviridae but were not able to create a proper alignment from which to estimate aligment scores that would resemble the kmer-counts that we obtained with the clustering method.
+We tried to reproduce the analysis in a small family such as Papillomaviridae but were not able to create a proper alignment from which to estimate alignment scores that would resemble the kmer-counts that we obtained with the clustering method.
+* provide more information about problems with alignment
 
@@
 
-Color edges in a gradient according with length of overlap
+* we already use edge width for number of proteins (sample size)
+* graphs are already saturated with information
+* manually annotate graphs with edge labels for overlap lengths
+* Discussion - briefly talk about how graphs could be modified to present other information, provide example as supplementary figure (show one)
 

### 2021-12-08 paper/response.md
adjusted wording in first response

@@
 It is a true and unfortunate fact that many records in Genbank are misannotated, even for entries in the RefSeq database that have presumably been manually curated.
-Unfortunately, due to the scale of our analysis (all available virus genomes), it was not feasible to curate the entire data set of 12,609 genomes (451,228 annotated reading frames) in the time allotted to revise our manuscript.
+Due to the scale of our analysis (all available virus genomes), it was not feasible to curate the entire data set of 12,609 genomes (451,228 annotated reading frames).
 However, there are some quality control measures that can be automated.
@@
 For example, an overlap between reading frames that are shifted by `+2` should have a length that is not divisible by 3, with remainder 1, *i.e.*, length mod 3 = 1.
-Applying this filter to our data identified 5,241 (~1%) entries with misannotations - these entries were discarded from further analysis.
+Applying this filter to our data identified 5,241 (~1%) entries with inconsistent overlap lengths - these entries were discarded from further analysis.
 We manually determined that these cases were frequently caused by undocumented splicing or slippage.
-
 Thus, the cases of misannotation that the reviewer astutely identified have already been filtered out of our analysis, though those entries remained in the source data.
 Due to space constraints, this quality control step was not described in our original manuscript.
-We sincerely apologize for the confusion, and we have added this explanation to the revision.
+We sincerely apologize for this misunderstanding, and we have added this explanation to the revision.
 

### 2021-12-08 paper/response.md
More coments

@@
 
-3. Biases on $+0$ frame shift
+3. Biases on +0 frame shift
 
@@
 7562-11329 and join(7562-9970,9970-10047) (ribosomal frameshift at 9970)
-> leading to $+0$ frame overlaps 45-5684 and 7562-9970.
+> leading to +0 frame overlaps 45-5684 and 7562-9970.
 On the other hand, the alphavirus NC\_023812 only has two CDSs annotated 43-7467 and 7541..11269.
@@
 
-We agree that this is a significant limitation in the analysis (see above, removed +0 from results)
+We agree that this is a significant limitation in the analysis (see above, removed +0 from results).
+As kindly suggested by the revieer, when re-plotting our figures, we see a change in the distribution of overlapping lengths, particularly in RNA viruses.
 
@@
 
-Thank you for raising this issue.
-As noted in our response to an earlier point, cases that involved splices such as the example provided by the reviewer were excluded from our analysis.
+Thank you for raising this issue. A closer look into it lead us to conclude that such cases appear when we try to calculate overlaps in coding sequences with internal ribosomal frameshifts.
 
-This was the correct file. How did we deal with those cases? Removing overlaps product from splicing events.
-The input for our calculation of overlaps is based on the file total\_orfs.csv that contains for each protein from each genome the accession number (accno), Product, Strand, Coordinates and Start Codon.
-The misscalculation problem seems to be associated with entries that have internal frame shifts.
-For example, for the record NC\_003092, the annotations for ORFS1ab protein and ORF1aTF polyprotein have some internal frame shifts that are not even spliced events: 209:6521;6520:10996 and 209:2876;2874:3552 respectively.
-Similarly, for NC\_001449 the truncated polyprotein has coordinates: 7561:9970;9969:10047 (again, a frame shift of one nucleotide) where ribosome probably jumps back one nucleotide.
-Calculating frame shift on such entries is challenging since there would be two different frame shifts for the same protein.
+For example, "ORFS1ab protein" and "ORF1aTF polyprotein" have the following coordinates:  209:6521;6520:10996 and 209:2876;2874:3552 respectively.
 
-Following step: Detect how many entries we have with similar characteristics.
+When calculating the overlaps of such reading frames, we are storing different overlap lengths for the different parts of the CDS as independent entries.
+We recognized a total of 1,426 these cases, and merge them to have one single entry per CDS with the length of the total overlap.
+
+However, as the reviewer pointed out, many of such cases are present on +0 frameshifts. Therefore, they where removed from the subsequent analysis.
 
@@
 
-## Minor Issues}
+## Minor Issues
 

### 2021-12-09 paper/response.md
updated response.md, but please use version on Overleaf!

--- 2021-12-08 paper/response.md
@@
-# Major Issues
+## Major Issues
 
-## Reviewer 1
+### Reviewer 1
 
-1. Annotation reliability
-
->The authors rely on NCBI annotations of CDSs.
+> 1) The authors rely on NCBI annotations of CDSs.
 However, these tend to be quite inaccurate (as the authors discovered for bovine adenovirus NC 002513, p20 lines 5-8).
@@
 Thank you for raising this issue.
-It is a true and unfortunate fact that many records in Genbank are misannotated, even for entries in the RefSeq database that have presumably been manually curated.
+It is a true and unfortunate fact that many records in Genbank are misannotated, even for the reference genomes used in our study, which were manually curated as an output of the NCBI Viral genomes Resource.
 Due to the scale of our analysis (all available virus genomes), it was not feasible to curate the entire data set of 12,609 genomes (451,228 annotated reading frames).
@@
 Next, we re-ran the analysis pipeline on the modified data to assess the impact of this error variation on our results.
+**IN PROGRESS**
 
@@
 
-2. Overlaps in the $+0$ frame shift
-
-> The authors include +0 frame overlaps.
+> 2) The authors include +0 frame overlaps.
 By this, they mean where the same coding sequence forms part of two different CDS annotations, e.g.
@@
 Therefore, we are providing supplementary material that includes the +0 cases so that this information is available, with the caveats raised by the reviewer.
+**IN PROGRESS**
 
-3. Biases on $+0$ frame shift
-
-> A second problem with including +0 frame overlaps is that their presence in a genome is very much subject to the whim of the annotator.
+> 3) A second problem with including +0 frame overlaps is that their presence in a genome is very much subject to the whim of the annotator.
 As an example, the alphavirus NC\_001449 has both readthrough and frameshift ORFs annotated:
@@
 
-We agree that this is a significant limitation in the analysis (see above, removed +0 from results)
+We agree that inconsistent annotation of +0 frame overlaps was a significant issue in our original analysis.
+As noted above, we have removed the +0 overlaps from our main results in the manuscript, and relegated the previous versions in Supplementary Material.
+**DONE**
 
-4. Annotation bug causing wrong overlap measurement
-
-> There seem to be issues with the annotation scripts.
+> 4) There seem to be issues with the annotation scripts.
 For example in the authors' supplementary overlaps.csv file, we have the following entries for NC\_001449, listing the +0 frame P123/P1234 overlap in the first line, but then the frameshift ORF join(7562-9970,9970-10047), comprising a +0 frame overlap 7562-9970 and a +2 frame overlap 9970-10047, is listed as two +0 frame overlaps:
@@
 As noted in our response to an earlier point, cases that involved splices such as the example provided by the reviewer were excluded from our analysis.
+**IN PROGRESS?**
 
@@
 
-## Reviewer #2
+### Reviewer #2
 
-1. Consistency of alignment-free method
-> In the graph analysis, homology is based on an alignment free method.
+> 1.
+In the graph analysis, homology is based on an alignment free method.
 how robust are the results of the graph analyses?
@@
 Such alignments tend to be affected for large periods of evolution that in viruses usually involve events such as recombination, insertion and deletions, and the lack of entire genes in different species.
-We tried to reproduce the analysis in a small family such as Papillomaviridae but were not able to create a proper alignment from which to estimate alignment scores that would resemble the kmer-counts that we obtained with the clustering method.
-* provide more information about problems with alignment
+An alternative approach would be to use pairwise alignment to generate a similar distance matrix for all coding sequences in a virus family.
+Since alignment is more computationally complex than matching k-mers, we wrote a message-passing interface (MPI)-enabled Python script to run the alignment program MAFFT in parallel (15 cores) and compute p-distances (*i.e.*, the proportion of aligned sites that are different and ungapped).
+It was then straightforward to graft the resulting distance matrix into our analysis pipeline.
+In the revised manuscript, we provide a set of clustering results for Papillomaviridae using pairwise alignment instead of the k-mer-based method.
+**IN PROGRESS**
 
-2. Overlap length information on graphs
-
-> The graph analysis is lacking information of the overlap frame and overlap length, which are of great interest.
+> 2.
+The graph analysis is lacking information of the overlap frame and overlap length, which are of great interest.
 Adding edges separate for frame and for length in the supplementary information would be of value (or at least adjusting edge width by a minimum of overlap length and discussing the impact on the resulting graph).
 
-* we already use edge width for number of proteins (sample size)
-* graphs are already saturated with information
-* manually annotate graphs with edge labels for overlap lengths
+We concur that our current layout algorithm for drawing adjacency graphs does not visualize information in the frame or length of overlaps.
+The key challenge is that there are limited channels for visually communicating information at our disposal, such as node size, edge width, node colour.
+We have avoided overloading the graph by, for example, varying both node shape, size and colour.
+As suggested by the reviewer, we revised our Python script for generating graph layouts, so that we can now provide examples of scaling edge widths to overlap lengths instead of the number of overlaps as supplementary material.
+
 * Discussion - briefly talk about how graphs could be modified to present other information, provide example as supplementary figure (show one)
 
-## Minor Issues}
+## Minor Issues
 
@@
 
-1. Suggested references
 > While the authors have referenced previous studies such as Brandes & Linial, Chirico et al, and Schlub & Holmes, I feel that they should also reference Rancurel et al PMID 19640978 and Pavesi et al PMID 30339683, who used carefully curated datasets of overlaps rather than relying on NCBI annotation.
 
-Include them
+Thank you for bringing these articles to our attention.
+We have incorporated these references in discussing the trade-off between accuracy and scaleability, with respect to the careful manual curation versus automated computational processing of virus genomes.
 
-2. Specify terms
 > The authors frequently (e.g.
@@
 
-3. Negative strand
-
 > On p6 lines 10-11, I feel it would be useful to state explicitly how (-)strand ORFs (where the biological 5' and 3' ends are in the opposite orientation with respect to the genome coordinates) are dealt with.
-
-4. Coronaviridae genome
 
@@
 Proteins with high homology to certain clussters can be grouped in that specific cluster and won't form their individual cluster.
-
-5. Simian hemorrhagic fever genome
 
@@
 
-1. Figures discrepancy
-
 > There seems to be a discrepancy between figure 1c and figure 2a - in overlap of frame +2 the majority of overlap lengths is 1 and 4, and overlap of +2 is the dominant in both dsDNA and dsRNA, however in Figure 1c there are no peaks for dsDNA with length 1 and 4, this does not make sense because there are more dsDNA than dsRNA in the analysis.
@@
 
-2. Overlaps on -0 frame shift
-
 > Long overlaps with -0 frame are not explained - is this unique to a single edge between homology clusters or does this occur in several edges, if the latter is true this might have implications for the evolution of the genetic code.
 
-3. Conserved overlaps
+> Although most adjacent nodes don't show overlap, some large nodes have thick edges (e.g., nodes 6 and 7 in the Adenoviridae family) suggesting that in some cases adjacency of specific genes is accompanied by conserved overlap between the two genes - this should be discussed.
 
-> Although most adjacent nodes don't show overlap, some large nodes have thick edges (e.g., nodes 6 and 7 in the Adenoviridae family) suggesting that in some cases adjacency of specific genes is accompanied by conserved overlap between the two genes - this should be discussed.
+Thank you for raising this point.  We have a discussion of this example in the revised manuscript.

### 2021-12-09 paper/response.md

@@
 
-This was the correct file. How did we deal with those cases? Removing overlaps product from splicing events.
+<!--This was the correct file. How did we deal with those cases? Removing overlaps product from splicing events.
 The input for our calculation of overlaps is based on the file total\_orfs.csv that contains for each protein from each genome the accession number (accno), Product, Strand, Coordinates and Start Codon.
@@
 Following step: Detect how many entries we have with similar characteristics.
+-->
 
@@
 
-* Discussion - briefly talk about how graphs could be modified to present other information, provide example as supplementary figure (show one)
+**ADDED TO DISCUSSION - just need supplementary figure**
 
@@
 We have incorporated these references in discussing the trade-off between accuracy and scaleability, with respect to the careful manual curation versus automated computational processing of virus genomes.
+**DONE**
 
@@
 
-change them
+Thank you for pointing this out.
+We agree that making the distinction between an "open reading frame" and a "reading frame" resolves the ambiguity of whether we are referring to a coding sequence or a context.
+We have replaced cases of the former with "ORF" (after first usage of "open reading frame"), and retain "reading frame" for the latter.
+**DONE**
 
@@
 
+**IN PROGRESS**
+
 > In Fig 4, why are there only 11 clusters in the coronaviridae graph, given that SARS alone has 13 ORFs and there are additional coronavirus accessory genes present in other coronaviruses?
 
-Proteins with high homology to certain clussters can be grouped in that specific cluster and won't form their individual cluster.
+Proteins with high homology to certain clusters can be grouped in that specific cluster and won't form their individual cluster.
+**IN PROGRESS**
 
@@
 
-Double check it
+We sincerely apologize for this mistake.
+The reviewer is absolutely correct - there are only 15 annotated coding sequences in this Genbank record.
+Our CSV file of ORFs contains the same number, and we recorded 13 overlaps, of which 5 involved `+0` frameshifts that we now discard from our revised analysis.
+The remaining 8 overlaps involved 10 ORFs.
+We have corrected this line accordingly.
+**DONE**
 
@@
 
+We apologize for causing confusion.
+Figure 1 is an analysis performed at the level of virus genomes, whereas Figure 2 is an analysis at the level of individual overlapping open reading frames (OvRFs).
+The apparent discrepancy described by the reviewer was the result of overlaps of a single nucleotide being enormously abundant in dsDNA genomes.
+For each dsDNA genome, this resulted in a single observed mean length close to 1 (Figure 1C), but many data points in the ridgeplot for frameshift +2 (Figure 2, left).
+We have added an explanation to the text, and corrected the x-axis label for Figure 2, which was erroneously labelled "Mean overlap length" instead of "Overlap length".
+
+**FIXED IN TEXT**
+**JUST NEED TO CHANGE FIGURE 2 axis label from "Mean overlap length" to "Overlap length"**
+
 > Long overlaps with -0 frame are not explained - is this unique to a single edge between homology clusters or does this occur in several edges, if the latter is true this might have implications for the evolution of the genetic code.
+
+We agree that this is an interesting pattern in the data.
+In our revised Discussion section, we refer to a recent observation by Lèbre and Gascuel (2017, J Theor Biol) that any combination of residues can be encoded without necessarily inducing a stop codon in the `-0` reading frame, due to redundancy in the genetic code.
+**DONE**
 
@@
 
-Thank you for raising this point.  We have a discussion of this example in the revised manuscript.
+Thank you for raising this point.  We have added a discussion of this example in the revised manuscript.
+**DONE**

### 2021-12-10 dataset_2020/df_inspection.R
Re-plotting without cero overlaps

@@ -8,8 +8,8 @@ require(grid)
 # Plot frame shift distribution
 ########################################################
 setwd('/home/lmunoz/Projects/ovrf-review/dataset_2020/')
-#overlaps <- read.csv('2_overlapping_march2020.csv', colClasses='character')
-overlaps <- read.csv('overlaps_with_noise.csv', colClasses='character')
+overlaps <- read.csv('2_overlapping_march2020.csv', colClasses='character')
+#overlaps <- read.csv('overlaps_with_noise_50.csv', colClasses='character')
 
 overlaps$extreme_left1 <- as.integer(overlaps$extreme_left1)
 overlaps$extreme_left2 <- as.integer(overlaps$extreme_left2)
@@ -40,6 +40,9 @@ circular <- which(non_splicing_overlaps$extreme_left1==non_splicing_overlaps$ext
 
 non_splicing_overlaps <- non_splicing_overlaps[-circular,]
 
+# Remove +0 frame shifts
+non_splicing_overlaps <- non_splicing_overlaps %>% filter(shift != "+0")
+
 overlaps <- non_splicing_overlaps
 
 # Ridgeplot
@@ -50,22 +53,90 @@ plot <- ggplot(non_splicing_overlaps, aes(x = log10(overlap), y = shift, group =
   theme_bw()
 plot
 
-#palete<-c("#32621f", "#769845", "#cacfa6", "#f1dba2", "#de8f42", "#c1462b")
-# +2, +1, +0, +1, -1, -2
-
-# Introduce Noise: Randomly select rows and change coordinates 
-n<-3
-modified_ov <- overlaps %>% 
-                        slice_sample(n=n)
-
+########################################################
+# Import and merge data
+########################################################
 colnames(non_splicing_overlaps)[1] <- "Accession"
-colnames(virus)[3] <- "Accession"
+setwd('/home/lmunoz/Projects/ovrf-review/dataset_2020/')
+overlaps <- non_splicing_overlaps
+orfs <- read.csv('total_orfs.csv')
+#orfs <- read.csv('total_orfs_50.csv') # with noise
+virus <- read.csv('try_baltimore_class.csv')
+colnames(virus)[2] <- "Accession"
+
+virus$Proteins[virus$Proteins=='-'] <- NA
+virus$Proteins <- as.integer(as.character(
+  virus$Proteins))
+summary(virus)
+
[…]

### 2021-12-10 dataset_2020/randomnoise.py

@@ -6,10 +6,16 @@ import csv
 import random
 
 source_path = Path("total_orfs.csv")
-target_path = Path(source_path.stem + "_1").with_suffix('.csv')
+target_path = Path(source_path.stem + "_50").with_suffix('.csv')
 
 def randcoord(coordinate):
-    return str(random.randrange(coordinate+1))
+    ran_value = None
+    if coordinate > 11:
+        ran_value = random.randrange(-10,10)
+    else:
+        ran_value = random.randrange(1,10)
+    
+    return str(ran_value + coordinate)
 
 def to_multi_coords(multi_coords_str):
     return [coords.split(":") for coords in multi_coords_str.split(";")]
@@ -32,7 +38,7 @@ with source_path.open() as source_file:
         writer = csv.DictWriter(target_file, columns)
         writer.writeheader()
         n_rows = 451228
-        percentage = 0.1
+        percentage = 0.5
         candidates = dict.fromkeys(random.sample(range(1, n_rows), int(n_rows*percentage)))
         for i, row in enumerate(reader):
             if i % 10000 == 0:

### 2021-12-10 paper_plots/plots_for_paper_2020.R

@@ -1,0 +1,206 @@
+library(RColorBrewer)
+library(ggplot2)
+options(scipen=999)
+library(plyr)
+library(dplyr)
+
+######################################################
+# Preparing information and color palette 
+######################################################
+setwd('/home/lmunoz/Projects/ovrf-review/dataset_2020')
+virus <- read.csv('no_cero_out_virus_df_inspection.csv', stringsAsFactors = F)
+#virus <- read.csv('50_NOISE_no_cero_out_virus_df_inspection.csv', stringsAsFactors = F)
+
+# DO NOT USE, PLOT IS SKEWED.
+# If NA, the virus do not have overlapping proteins, therefore == 0 
+virus[which(is.na(virus$n.overlaps)),]$n.overlaps <- 0
+#virus[which(is.na(virus$len.overlaps)),]$len.overlaps <- 0
+#virus[which(is.na(virus$total.overlaps)),]$total.overlaps <- 0
+
+#Re naming Unknown according to molecule type
+virus$baltimore.class[which(virus$baltimore.class == "Unknown" & grepl("RNA", virus$Molecule))] <- "unknown_RNA"
+virus$baltimore.class[which(virus$baltimore.class == "Unknown" & grepl("DNA", virus$Molecule))] <- "unknown_DNA"
+
+# Setting color palette
+bal.class <- c("ds_RNA", "ss_RNA_+", "ss_RNA_-", "unknown_RNA", "ds_DNA", "ss_DNA", "unknown_DNA", "RT_viruses", "circular_ss_RNA")
+col<- c("#1d6996", "#38a6a5", "#57a1b4", "#87bfdb", "#edad08", "#e17c05", "#cc503e", "#808080", "#73af48")
+shape <- c(8, 4, 6, 5, 17, 18, 20, 1, 4)
+pal <- as.data.frame(cbind(bal.class, col))
+sh <- cbind.data.frame(bal.class, shape)
+
+#######################################################
+# Barplot database distribution
+#######################################################
+tbl <- as.data.frame(table(virus$baltimore.class))
+tbl$bal.class <- factor(tbl$Var1, levels=c("ds_DNA", "ss_DNA", "unknown_DNA", "ds_RNA", "ss_RNA_+", "ss_RNA_-", "unknown_RNA", "RT_viruses" ))
[…]
